Defending Against Manipulation: The Impact of Psychological Inoculation on Social Media Discernment

### 2024-12-01 paper/paper.qmd
Updated scaling

---
title: "Defending Against Manipulation: The Impact of Psychological Inoculation on Social Media Discernment"
subtitle: "An Analysis of Technique Discernment and Demographic Influences"
author: 
  - Elizabeth Luong
thanks: "Special thanks to Rohan Alexander for his guidance. Data and code are available at https://github.com/Luongel1/psychological-inoculation.git."
date: today
date-format: long
abstract: "Social media platforms have increasingly become battlegrounds for the dissemination of manipulative and deceptive content. This study investigates the efficacy of psychological inoculation in enhancing users' ability to discern manipulative from neutral content on social media, a concept referred to as 'technique discernment.' Utilizing Bayesian regression models and survey data, we examine the relationship between inoculation treatments and discernment scores while accounting for demographic and psychological covariates. The results reveal significant differences in discernment ability between treatment and control groups, highlighting the promise of inoculation as a tool for combating misinformation. These findings underline the importance of psychological resilience in navigating online ecosystems and inform strategies for future interventions."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(here)
library(rstanarm)
library(modelsummary)
library(ggplot2)
library(kableExtra)
library(arrow)
library(gridExtra)
library(knitr)

@@

The final cleaned dataset included the following variables:
`treatment`, `technique_discernment`, `standardized_analytical_thinking`, `standardized_conspiracy_belief`, `standardized_bullshit_receptivity`, `gender`, `education`, `political_ideology`, and `age_text.`

The cleaned dataset was summarized to provide an overview of its structure and distribution. @tbl-clean-data provides a preview of the first five rows of the dataset, highlighting key variables across treatment groups, psychological measures, and demographics.

```{r}
#| label: tbl-clean-data
#| tbl-cap: "Preview of the cleaned OSF dataset"
#| echo: false
#| warning: false
#| message: false

# Preview of the first few rows of the dataset
cleaned_data %>%
  slice_head(n = 5) %>%
  select(treatment, technique_discernment, standardized_analytical_thinking,
         standardized_conspiracy_belief, standardized_bullshit_receptivity,
         gender, education, political_ideology, age_text) %>%
  kable(booktabs = TRUE, linesep = "") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE) %>%
  add_header_above(c(" " = 3, "Psychological Variables" = 3, "Demographics" = 3))

@@

# Summary statistics for numeric variables
numeric_summary <- cleaned_data %>%
  select(technique_discernment, standardized_analytical_thinking,
         standardized_conspiracy_belief, standardized_bullshit_receptivity) %>%
  summary()

numeric_summary %>%
  kable(booktabs = TRUE, linesep = "") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Numeric Variables Summary" = 4))
```

@@

# Load necessary libraries
if (!require("knitr")) install.packages("knitr")
if (!require("kableExtra")) install.packages("kableExtra")
library(dplyr)
library(knitr)
library(kableExtra)

@@

# Recalculate group indices for clear grouping
group_indices <- categorical_summary %>%
  group_by(Variable) %>%
  summarize(
    start = min(row_number()),
    end = max(row_number()),
    .groups = "drop"
  )

# Create a well-formatted table using kableExtra
table_output <- categorical_summary %>%
  kbl(
    booktabs = TRUE,
    col.names = c("Variable", "Category", "Frequency"),
    caption = "Summary of Categorical Variables"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center",
    font_size = 14
  )

# Add group rows dynamically for better alignment and clarity
for (i in seq_len(nrow(group_indices))) {
  table_output <- table_output %>%
    group_rows(group_indices$Variable[i], group_indices$start[i], group_indices$end[i])
}

# Display the final table
table_output

@@

@tbl-categorical demonstrates a balanced experimental design between the treatment groups, with 2,757 participants in the Control group and 2,748 in the Inoculation group. Gender distribution reflects a higher representation of females (2,751) and males (2,685), alongside smaller groups of non-binary individuals (52) and those preferring not to disclose their gender (7). Regarding education, the data is skewed towards participants with advanced degrees, as evidenced by 2,306 participants holding Doctorates and 1,709 with Master's degrees, whereas only a small fraction reported High School (37) or Bachelor's education (149). Political ideology distribution shows considerable diversity: 2,201 participants identify as Liberal, 2,147 as Conservative, and 1,157 as Moderate. Age distribution spans a broad range, with the largest cohort being participants aged 55 or older (1,528) and the smallest being those aged 18–24 (707).

@@

The standardized psychological variables: `standardized_analytical_thinking`, `standardized_bullshit_receptivity`, and `standardized_conspiracy_belief`—are key metrics for evaluating cognitive and emotional responses to misinformation. @fig-psych compares these variables across treatment groups, revealing slight but consistent differences favoring the inoculation group. For instance, the inoculation group shows lower susceptibility to conspiracy beliefs, aligning with the intended effects of the intervention.

@@

# Reshape data to long format
long_data <- cleaned_data %>%
  pivot_longer(cols = c(standardized_analytical_thinking, 
                        standardized_bullshit_receptivity, 
                        standardized_conspiracy_belief),
               names_to = "variable",
               values_to = "value")

# Plot using facets
ggplot(long_data, aes(x = treatment, y = value, fill = treatment)) +
  geom_violin(alpha = 0.7) +
  labs(title = "Psychological Variables by Treatment", 
       x = "Treatment Group", 
       y = "Standardized Value") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  facet_wrap(~ variable, scales = "free_y")

@@

# Ensure required libraries are installed
if (!require("modelsummary")) install.packages("modelsummary")
if (!require("broom.mixed")) install.packages("broom.mixed")
if (!require("rstanarm")) install.packages("rstanarm")

# Load required libraries
library(modelsummary)
library(broom.mixed)
library(rstanarm)

@@

# Display the table using `kableExtra`
library(knitr)
library(kableExtra)

@@

# Ensure required libraries are installed
if (!require("rstanarm")) install.packages("rstanarm")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("dplyr")) install.packages("dplyr")

# Load libraries
library(rstanarm)
library(ggplot2)
library(dplyr)

@@

@@

model <- stan_glm(
  formula = technique_discernment ~ treatment + standardized_analytical_thinking + 
    standardized_conspiracy_belief + standardized_bullshit_receptivity + 
    gender + education + political_ideology + age_text,
  data = cleaned_data,
  family = gaussian(),
  prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
  prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
  prior_aux = exponential(rate = 1, autoscale = TRUE),
  seed = 853
)

@@

```{r}
#| label: fig-trace-plots
#| fig-cap: "Trace plots for key model parameters"
#| fig-subcap: ["Intercept", "Treatment: Inoculation", "Standardized Analytical Thinking"]
#| layout-ncol: 3
#| echo: false
#| eval: true
#| warning: false
#| message: false

plot(model, "trace", pars = c("(Intercept)", "treatmentInoculation", "standardized_analytical_thinking"))

@@

plot(model, "rhat")

@@

# Load the required library
if (!require("car")) install.packages("car")
library(car)

@@

vif_values %>%
  as.data.frame() %>%
  rownames_to_column("Predictor") %>%
  kable(
    booktabs = TRUE,
    caption = "VIF values for Bayesian regression predictors"
  ) %>%
  kable_styling(full_width = FALSE, position = "center")

```
## Credibility Intervals
The 90% credibility intervals for the model coefficients are visualized in @fig-credibility-intervals. These intervals highlight the uncertainty around the estimated effects for each predictor. Notably, the treatmentInoculation variable has a positive and significant effect, providing evidence for the effectiveness of psychological inoculation interventions.

@@

model_summary %>%
  dplyr::select(term, estimate, std.error, conf.low, conf.high) %>%
  dplyr::mutate(
    term = stringr::str_replace_all(term, "_", " "), # Format terms for readability
    conf.int = paste0("(", round(conf.low, 2), ", ", round(conf.high, 2), ")")
  ) %>%
  dplyr::select(term, estimate, std.error, conf.int) %>%
  kbl(
    caption = "Detailed Bayesian regression model summary",
    col.names = c("Predictor", "Estimate", "Standard Error", "95% CI")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )

>>>>

---
title: "Defending Against Manipulation: The Impact of Psychological Inoculation on Social Media Discernment"
subtitle: "An Analysis of Technique Discernment and Demographic Influences"
author: 
  - Elizabeth Luong
thanks: "Special thanks to Rohan Alexander for his guidance. Data and code are available at https://github.com/Luongel1/psychological-inoculation.git."
date: today
date-format: long
abstract: "Social media platforms have increasingly become battlegrounds for the dissemination of manipulative and deceptive content. This study investigates the efficacy of psychological inoculation in enhancing users' ability to discern manipulative from neutral content on social media, a concept referred to as 'technique discernment.' Utilizing Bayesian regression models and survey data, we examine the relationship between inoculation treatments and discernment scores while accounting for demographic and psychological covariates. The results reveal significant differences in discernment ability between treatment and control groups, highlighting the promise of inoculation as a tool for combating misinformation. These findings underline the importance of psychological resilience in navigating online ecosystems and inform strategies for future interventions."
output: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(here)
library(rstanarm)
library(modelsummary)
library(ggplot2)
library(kableExtra)
library(arrow)
library(gridExtra)
library(knitr)
library(car)
library(dplyr)
library(stringr)
library(magrittr)
library(tibble)
library(bayesplot)
library(broom.mixed)

@@

The final cleaned dataset included the following variables:
`treatment`, `technique_discernment`, `standardized_analytical_thinking`,

`standardized_conspiracy_belief`,
`standardized_bullshit_receptivity`, `gender`, `education`, `political_ideology`, and `age_text.`

The cleaned dataset was summarized to provide an overview of its structure and distribution. @tbl-clean-data provides a preview of the first five rows of the dataset, highlighting key variables across treatment groups, psychological measures, and demographics.
```{r}
#| label: tbl-clean-data
#| tbl-cap: "Preview of the cleaned OSF dataset"
#| echo: false
#| warning: false
#| message: false

# Preview of the first few rows of the dataset with cleaned column names
cleaned_data %>%
  slice_head(n = 5) %>%
  select(
    treatment, technique_discernment, standardized_analytical_thinking,
    standardized_conspiracy_belief, standardized_bullshit_receptivity,
    gender, education, political_ideology, age_text
  ) %>%
  rename_with(~ stringr::str_replace_all(., "_", " ")) %>% # Replace underscores with spaces in column names
  kable(booktabs = TRUE, linesep = "", longtable = TRUE) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    font_size = 7, # Slightly reduce font size for better fit
    position = "center"
  ) %>%
  add_header_above(
    c(" " = 1, "Psychological Variables" = 4, "Demographics" = 4) # Correct grouping for all 9 columns
  ) %>%
   column_spec(1, width = "1.8cm", bold = TRUE) %>% # Adjust column widths for clarity
  column_spec(2:5, width = "1.5cm") %>% # Narrower columns for Psychological Variables
  column_spec(6:9, width = "1.8cm") %>% # Adjust for Demographics
  row_spec(0, bold = TRUE, background = "#f7f7f7") # Header row styling

@@

# Summary statistics for numeric variables
numeric_summary <- cleaned_data %>%
  select(
    technique_discernment,
    standardized_analytical_thinking,
    standardized_conspiracy_belief,
    standardized_bullshit_receptivity
  ) %>%
  summary()

# Rename variables to replace underscores with spaces
colnames(numeric_summary) <- stringr::str_replace_all(
  colnames(numeric_summary),
  "_",
  " "
)

# Render the table with adjusted formatting
numeric_summary %>%
  kable(
    booktabs = TRUE,
    linesep = "",
    longtable = TRUE, # Enables table spanning multiple pages
    align = "c" # Center-align contents
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = TRUE, # Ensures the table doesn't stretch
    font_size = 10, # Adjust font size for better readability
    position = "center" # Centers the table on the page
  ) %>%
  add_header_above(
    c(" " = 1, "Numeric Variables Summary" = 4) # Correct grouping for the header
  ) %>%
  column_spec(1, bold = TRUE, width = "4cm") %>% # Adjust and bold the first column
  column_spec(2:5, width = "3cm") %>% # Adjust widths for numeric columns
  row_spec(0, bold = TRUE, background = "#f7f7f7") # Style the header row

@@

@@

# Format table to show variable headers with their respective details listed beneath
formatted_table <- categorical_summary %>%
  group_by(Variable) %>%
  mutate(
    Variable = if_else(row_number() == 1, Variable, "") # Show Variable name only once
  ) %>%
  ungroup() %>%
  select(Variable, Category, Frequency) # Select the relevant columns

# Create a well-formatted hierarchical table using kableExtra
formatted_table %>%
  kbl(
    booktabs = TRUE,
    col.names = c("Variable", "Category", "Frequency"),
    caption = "Summary of Categorical Variables"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center",
    font_size = 12
  ) %>%
  column_spec(1, bold = TRUE, width = "10em") %>% # Bold and adjust width for Variable column
  column_spec(2, width = "12em") %>% # Adjust width of Category column
  column_spec(3, width = "8em") %>% # Adjust width of Frequency column
  row_spec(0, bold = TRUE, background = "#f2f2f2") # Style header row

@@

@tbl-categorial demonstrates a balanced experimental design between the treatment groups, with 2,757 participants in the Control group and 2,748 in the Inoculation group. Gender distribution reflects a higher representation of females (2,751) and males (2,685), alongside smaller groups of non-binary individuals (52) and those preferring not to disclose their gender (7). Regarding education, the data is skewed towards participants with advanced degrees, as evidenced by 2,306 participants holding Doctorates and 1,709 with Master's degrees, whereas only a small fraction reported High School (37) or Bachelor's education (149). Political ideology distribution shows considerable diversity: 2,201 participants identify as Liberal, 2,147 as Conservative, and 1,157 as Moderate. Age distribution spans a broad range, with the largest cohort being participants aged 55 or older (1,528) and the smallest being those aged 18–24 (707).

@@

The standardized psychological variables: `standardized_analytical_thinking`,

`standardized_bullshit_receptivity`, and `standardized_conspiracy_belief`—are key metrics for evaluating cognitive and emotional responses to misinformation. @fig-psych compares these variables across treatment groups, revealing slight but consistent differences favoring the inoculation group. For instance, the inoculation group shows lower susceptibility to conspiracy beliefs, aligning with the intended effects of the intervention.

@@

# Reshape data to long format
long_data <- cleaned_data %>%
  pivot_longer(
    cols = c(standardized_analytical_thinking, 
             standardized_bullshit_receptivity, 
             standardized_conspiracy_belief),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(variable = stringr::str_replace_all(variable, "_", " ")) # Replace underscores with spaces

# Plot using facets
ggplot(long_data, aes(x = treatment, y = value, fill = treatment)) +
  geom_violin(alpha = 0.7) +
  labs(
    title = "Psychological Variables by Treatment", 
    x = "Treatment Group", 
    y = "Standardized Value"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  facet_wrap(~ variable, scales = "free_y")

@@

@@

@@

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

model <- stan_glm(
  formula = technique_discernment ~ treatment + standardized_analytical_thinking + 
    standardized_conspiracy_belief + standardized_bullshit_receptivity + 
    gender + education + political_ideology + age_text,
  data = cleaned_data,
  family = gaussian(),
  prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
  prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
  prior_aux = exponential(rate = 1, autoscale = TRUE),
  seed = 853
)
```

@@

@@

```{r}
#| label: fig-trace-plots
#| fig-cap: "Trace plots for key model parameters"
#| fig-subcap: ["Intercept", "Treatment: Inoculation", "Standardized Analytical Thinking"]
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Create trace plots for specific parameters
trace_plot <- mcmc_trace(
  as.array(model), 
  pars = c("(Intercept)", "treatmentInoculation", "standardized_analytical_thinking"),
  facet_args = list(ncol = 1) # Arrange plots vertically
)

# Replace "_" with " " in parameter titles
trace_plot <- trace_plot +
  scale_y_discrete(labels = c(
    "(Intercept)" = "Baseline: Intercept",
    "treatmentInoculation" = "Treatment: Inoculation",
    "standardized_analytical_thinking" = "Standardized Analytical Thinking"
  )) +
  theme_minimal(base_size = 5) +
  labs(
    title = "Trace Plots for Key Model Parameters",
    x = "Iteration",
    y = "Parameter Values"
  ) +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    strip.text = element_text(size = 14)
  )

# Display the trace plot
trace_plot

@@

# Extract Rhat values
rhat_values <- bayesplot::rhat(model)  # This works directly with a fitted Bayesian model object

# Convert Rhat values to a data frame
rhat_data <- as.data.frame(rhat_values) %>%
  rownames_to_column(var = "Parameter") %>%
  rename(Rhat = rhat_values) %>%
  mutate(
    Parameter = gsub("_", " ", Parameter), # Replace underscores with spaces
    Parameter = stringr::str_replace_all(Parameter, "\\(Intercept\\)", "Baseline: Intercept"),
    Parameter = stringr::str_replace(Parameter, "treatmentInoculation", "treatment: Inoculation"),
    Parameter = stringr::str_replace(Parameter, "genderMale", "gender: Male"),
    Parameter = stringr::str_replace(Parameter, "genderNon-binary", "gender: Non-binary"),
    Parameter = stringr::str_replace(Parameter, "genderPrefer not to say", "gender: Prefer not to say"),
    Parameter = stringr::str_replace(Parameter, "educationDoctorate", "education: Doctorate"),
    Parameter = stringr::str_replace(Parameter, "educationHigh School", "education: High School"),
    Parameter = stringr::str_replace(Parameter, "educationMaster's", "education: Master's"),
    Parameter = stringr::str_replace(Parameter, "political ideologyLiberal", "political ideology: Liberal"),
    Parameter = stringr::str_replace(Parameter, "political ideologyModerate", "political ideology: Moderate"),
    Parameter = stringr::str_replace(Parameter, "age text25-34", "age: 25-34"),
    Parameter = stringr::str_replace(Parameter, "age text35-44", "age: 35-44"),
    Parameter = stringr::str_replace(Parameter, "age text45-54", "age: 45-54"),
    Parameter = stringr::str_replace(Parameter, "age text55 or older", "age: 55 or older")
  )

# Plot Rhat values
ggplot(rhat_data, aes(x = Rhat, y = reorder(Parameter, Rhat))) +
  geom_point(size = 3, color = "blue") +
  labs(
    title = "Rhat Diagnostic for Model Convergence",
    x = "Rhat Values",
    y = "Model Parameters"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  ) +
  geom_vline(xintercept = 1.1, linetype = "dashed", color = "red") # Highlight Rhat threshold

@@

@@

vif_values %>%
  as.data.frame() %>%
  rownames_to_column("Predictor") %>%
  mutate(Predictor = gsub("_", " ", Predictor)) %>% # Replace underscores with spaces
  kable(
    booktabs = TRUE,
    caption = "VIF values for Bayesian regression predictors"
  ) %>%
  kable_styling(full_width = FALSE, position = "center")
```
## Credibility Intervals
The 90% credibility intervals for the model coefficients are visualized in @fig-credibility-intervals. These intervals highlight the uncertainty around the estimated effects for each predictor. Notably, the treatmentInoculation variable has a positive and significant effect, providing evidence for the effectiveness of psychological inoculation interventions.

@@

# Update labels for predictors
model_summary <- model_summary %>%
  dplyr::mutate(
    term = stringr::str_replace_all(term, "_", " "), # Replace underscores with spaces
    term = stringr::str_replace_all(term, "\\(Intercept\\)", "Baseline: Intercept"), # Format intercept
    term = stringr::str_replace(term, "treatmentInoculation", "treatment: Inoculation"),
    term = stringr::str_replace(term, "genderMale", "gender: Male"), # Format gender terms
    term = stringr::str_replace(term, "genderNon-binary", "gender: Non-binary"),
    term = stringr::str_replace(term, "genderPrefer not to say", "gender: Prefer not to say"),
    term = stringr::str_replace(term, "educationDoctorate", "education: Doctorate"), # Format education terms
    term = stringr::str_replace(term, "educationHigh School", "education: High School"),
    term = stringr::str_replace(term, "educationMaster's", "education: Master's"),
    term = stringr::str_replace(term, "political ideologyLiberal", "political ideology: Liberal"), # Format ideology
    term = stringr::str_replace(term, "political ideologyModerate", "political ideology: Moderate"),
    term = stringr::str_replace(term, "age text25-34", "age: 25-34"), # Format age groups
    term = stringr::str_replace(term, "age text35-44", "age: 35-44"),
    term = stringr::str_replace(term, "age text45-54", "age: 45-54"),
    term = stringr::str_replace(term, "age text55 or older", "age: 55 or older")
  )

# Generate and display the table with formatted labels
model_summary %>%
  dplyr::select(term, estimate, std.error, conf.low, conf.high) %>%
  dplyr::mutate(
    conf.int = paste0("(", round(conf.low, 2), ", ", round(conf.high, 2), ")") # Format confidence intervals
  ) %>%
  dplyr::select(term, estimate, std.error, conf.int) %>%
  kbl(
    caption = "Detailed Bayesian regression model summary with formatted predictor labels",
    col.names = c("Predictor", "Estimate", "Standard Error", "95% CI")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )
```
